```{r all_flags}
#| echo: true
#| warning: false
#| message: false
#| collapse: true
```

### Key Features of the Dataset

Each row in the data set represents an individual subject and includes the following selected variables:

-   **Subject demographics**: age, sex, ethnicity, education level, income level, and employment status

-   **Relationship context**: whether the subject is part of a same-sex relationship, whether they are married, the duration of the relationship and how many children are in the household.

-   **Couple behavior indicators**: sex, flirting and fighting frequency.

-   **Pandemic Variables**: interpreted impact of the pandemic in the relationship, income change during the pandemic, if the subject and their partner were sick with COVID-19 and if they were vaccinated.

-   **Quality of the relationship**: variable used to measure the subject's perceived quality of their relationship.

### Purpose and Use Cases

This data set supports investigations into:

-   Demographic and behavioral predictors of relationship quality.

-   How the pandemic experience affected relationships.

-   Differences in relationship dynamics between different levels of income, sex and sexual orientation.

## Case Study

### Objective

**What behavioral factors are strongly associated with relationship quality in the context of the COVID-19 pandemic? Are these factors different for same sex couples?**

This case study examines the association between `relationship_quality` and a variety of behavioral and pandemic-related variables.

By examining survey data, we aim to:

-   Identify the behavioral factors that most strongly affect the perceived relationship quality. We will focus specifically on finding how `sex_frequency`, `flirts_with_partner` and `fights_with_partner` affect the perception of relationship quality and we are going to evaluate if these factors have a different impact on same sex couples (`same_sex_couple`).
-   We will also evaluate whether COVID-19 related factors such vaccination status (`subject_vaccinated`, `partner_vaccinated`), COVID-19 illness (`subject_had_covid`, `partner_had_covid`), income changes during the pandemic (`inc_change_during_pandemic`), and agreement on pandemic-related approaches (`agree_covid_approach`) affected couples' perceived relationship quality.

### Analysis

First, we need to load the libraries that will be used through the exercise.

```{r library_load, results = "hide", warning = FALSE, message = FALSE}
library(tidyverse) 
library(scales)
library(cowplot)
library(MASS)
library(broom)
library(brant)
library(knitr)
```

#### 1. Data Cleaning & Processing

First, we can load our data set, select our specific variables of intereset and remove NAs from our data set.

```{r data_load, warning = FALSE, message = FALSE}
# Reading Data
hcmst <- read_csv("../../data/clean/hcmst.csv") 

# Review total rows
nrow(hcmst)

# Select features of interest and removing NA
hcmst <- hcmst |> 
  dplyr::select(
    same_sex_couple, 
    sex_frequency, 
    flirts_with_partner, 
    fights_with_partner,
    inc_change_during_pandemic, 
    subject_had_covid, 
    partner_had_covid, 
    subject_vaccinated, 
    partner_vaccinated, 
    agree_covid_approach, 
    relationship_quality) |> 
  drop_na()

# Remaining row count.
nrow(hcmst)
```

#### 2. Variable Encoding

We can note that our categorical response follows a specific order. For this reason, we require a model that suits ordinal responses. Hence, we will be using an Ordinal Logistic regression model framework.

In order to do so, we need to make sure that our response variable is encoded as an ordered factor. We will also take this opportunity to encode our explanatory variables as factors (and ordered factors) as needed.

First, we can encode our nomical categorical explanatory variables:

```{r recode_a, warning = FALSE, message = FALSE}
hcmst <- hcmst |> 
  mutate(same_sex_couple = as_factor(same_sex_couple)|> 
           fct_recode("No" = "no", "Yes" = "yes"))

levels(hcmst$same_sex_couple)
```

Then, we can encode all our ordinal, categorical, explanatory, behavioral variables:

```{r recode_b}
hcmst <- hcmst |> 
  mutate(sex_frequency = as.ordered(sex_frequency)) |> 
  mutate(
    sex_frequency = fct_relevel(
      sex_frequency, 
      c(
        "once_a_month_or_less", 
        "2_to_3_times_a_month", 
        "once_or_twice_a_week", 
        "3_to_6_times_a_week", 
        "once_or_more_a_day"
        )
      ) |> 
      fct_recode(
        "Once a Month or Less"     = "once_a_month_or_less",
        "Two or Three Times a Month" = "2_to_3_times_a_month",
        "Once or Twice a Week"     = "once_or_twice_a_week",
        "Three to Six Times a Week" = "3_to_6_times_a_week",
        "Once or More a Day"       = "once_or_more_a_day"
      ))

levels(hcmst$sex_frequency)

hcmst <- hcmst |> 
  mutate(flirts_with_partner = as.ordered(flirts_with_partner)) |> 
  mutate(
    flirts_with_partner = fct_relevel(
      flirts_with_partner, 
      c(
        "never", 
        "less_than_once_a_month", 
        "1_to_3_times_a_month", 
        "once_a_week", 
        "a_few_times_a_week", 
        "every_day"
        )
      ) |> 
      fct_recode(
        "Never" = "never",
        "Less Than Once a Month" = "less_than_once_a_month",
        "One to Three Times a Month" = "1_to_3_times_a_month",
        "Once a Week" = "once_a_week",
        "A Few Times a Week" = "a_few_times_a_week",
        "Every Day" = "every_day"
        ))

levels(hcmst$flirts_with_partner)

hcmst <- hcmst |> 
  mutate(fights_with_partner = as.ordered(fights_with_partner)) |> 
  mutate(
    fights_with_partner = fct_relevel(
      fights_with_partner, 
      c(
        "0_times", 
        "1_time", 
        "2_times", 
        "3_times", 
        "4_times", 
        "5_times", 
        "6_times", 
        "7_or_more_times"
        )
      ) |> 
      fct_recode(
        "Zero Times" = "0_times",
        "One Time" = "1_time",
        "Two Times" = "2_times",
        "Three Times" = "3_times",
        "Four Times" = "4_times",
        "Five Times" = "5_times",
        "Six Times" = "6_times",
        "Seven or More Times" = "7_or_more_times"
      ))

levels(hcmst$sex_frequency)
```

Then, we can encode all our ordinal, categorical, explanatory, COVID-19-related variables:

```{r recode_c}
hcmst <- hcmst |> 
  mutate(inc_change_during_pandemic = as.ordered(inc_change_during_pandemic)) |> 
  mutate(
    inc_change_during_pandemic = fct_relevel(
      inc_change_during_pandemic, 
      c(
        "much_worse", 
        "worse", 
        "no_change", 
        "better", 
        "much_better"
        )
      ) |> 
      fct_recode(
        "Much Worse" = "much_worse",
        "Worse" = "worse",
        "No Change" = "no_change",
        "Better" = "better",
        "Much Better" = "much_better"
      ))

levels(hcmst$inc_change_during_pandemic)

hcmst <- hcmst |> 
  mutate(subject_had_covid = as_factor(subject_had_covid)|> 
           fct_recode("No" = "no", "Yes" = "yes"))

levels(hcmst$subject_had_covid)

hcmst <- hcmst |> 
  mutate(partner_had_covid = as_factor(partner_had_covid)|> 
           fct_recode("No" = "no", "Yes" = "yes"))

levels(hcmst$partner_had_covid)

hcmst <- hcmst |> 
  mutate(subject_vaccinated = as.ordered(subject_vaccinated)) |> 
  mutate(
    subject_vaccinated = fct_relevel(
      subject_vaccinated, 
      c(
        "not_vaccinated", 
        "partially_vaccinated", 
        "fully_vaccinated_no_booster", 
        "fully_vaccinated_and_booster"
        )
      ) |> 
      fct_recode(
        "Not Vaccinated" = "not_vaccinated",
        "Partially Vaccinated" = "partially_vaccinated",
        "Fully Vaccinated No Booster" = "fully_vaccinated_no_booster",
        "Fully Vaccinated and Booster" = "fully_vaccinated_and_booster"
      ))

levels(hcmst$subject_vaccinated)

hcmst <- hcmst |> 
  mutate(partner_vaccinated = as.ordered(partner_vaccinated)) |> 
  mutate(
    partner_vaccinated = fct_relevel(
      partner_vaccinated, 
      c(
        "not_vaccinated", 
        "partially_vaccinated", 
        "fully_vaccinated_no_booster", 
        "fully_vaccinated_and_booster"
        )
      ) |> 
      fct_recode(
        "Not Vaccinated" = "not_vaccinated",
        "Partially Vaccinated" = "partially_vaccinated",
        "Fully Vaccinated No Booster" = "fully_vaccinated_no_booster",
        "Fully Vaccinated and Booster" = "fully_vaccinated_and_booster"
      ))

levels(hcmst$partner_vaccinated)

hcmst <- hcmst |> 
  mutate(agree_covid_approach = as.ordered(agree_covid_approach)) |> 
  mutate(
    agree_covid_approach = fct_relevel(
      agree_covid_approach, 
      c(
        "completely_disagree", 
        "mostly_disagree", 
        "mostly_agree", 
        "completely_agree"
        )
      ) |> 
      fct_recode(
        "Completely Disagree" = "completely_disagree",
        "Mostly Disagree" = "mostly_disagree",
        "Mostly Agree" = "mostly_agree",
        "Completely Agree" = "completely_agree"
      ))

levels(hcmst$agree_covid_approach)
```

Finally, we can encode our target variable:

```{r recode_d}
hcmst <- hcmst |> 
  mutate(relationship_quality = as.ordered(relationship_quality)) |> 
  mutate(
    relationship_quality = fct_relevel(
      relationship_quality, 
      c(
        "very_poor", 
        "poor", 
        "fair", 
        "good", 
        "excellent"
        )
      ) |> 
      fct_recode(
        "Very Poor" = "very_poor",
        "Poor" = "poor",
        "Fair" = "fair",
        "Good" = "good",
        "Excellent" = "excellent"
      ))

levels(hcmst$sex_frequency)
```

#### 3. Exploratory Data Analysis

Let's visualize histograms charts for our different categorical variables to see how different groups were sampled in the survey and how different these samples are for same sex couples.

First, we will create a function that will be used for visualization.

```{r hist_function}
histogram_plot <- function(data, y_var, facet, plot_title = "Counts by Category") {

  data_long <- data |> 
    dplyr::select(all_of(c(y_var, facet))) |> 
    pivot_longer(cols = all_of(y_var), names_to = "var", values_to = "value") |> 
    count(var, value, facet_value = .data[[facet]])

  ggplot(data_long, aes(x = n, y = reorder(value, n), fill = facet_value)) +
    geom_bar(stat = "identity", position = "dodge", color = "white") +
    labs(
      title = plot_title,
      x = "Count",
      y = NULL,
      fill = "Same Sex Couple"
    ) +
    theme_minimal()
}
```

```{r histograms, fig.height=20, fig.width=8}

hist_sex_freq <- histogram_plot(
  hcmst, 
  "sex_frequency", 
  "same_sex_couple", 
  plot_title = "Sex Frequency"
)

hist_flirt_freq <- histogram_plot(
  hcmst, 
  "flirts_with_partner", 
  "same_sex_couple", 
  plot_title = "Flirting Frequency"
)

hist_fight_freq <- histogram_plot(
  hcmst, 
  "fights_with_partner", 
  "same_sex_couple", 
  plot_title = "Fighting Frequency"
)
    
hist_inc_change_freq <- histogram_plot(
  hcmst, 
  "inc_change_during_pandemic", 
  "same_sex_couple", 
  plot_title = "Income Change During the Pandemic Frequency"
)
    
hist_sub_covid_freq <- histogram_plot(
  hcmst, 
  "subject_had_covid", 
  "same_sex_couple", 
  plot_title = "Subject had COVID-19 Frequency"
)

hist_par_covid_freq <- histogram_plot(
  hcmst, 
  "partner_had_covid", 
  "same_sex_couple", 
  plot_title = "Subject's Partner had COVID-19 Frequency"
)

hist_sub_vax_freq <- histogram_plot(
  hcmst, 
  "subject_vaccinated", 
  "same_sex_couple", 
  plot_title = "Subject Vaccination Status Frequency"
)

hist_par_vax_freq <- histogram_plot(
  hcmst, 
  "partner_vaccinated", 
  "same_sex_couple", 
  plot_title = "Subject's Partner Vaccination Status Frequency"
)

hist_cov_approach_freq <- histogram_plot(
  hcmst, 
  "agree_covid_approach", 
  "same_sex_couple", 
  plot_title = "Couple's Agreement on Pandemic Approach Frequency"
)

hist_quality_freq <- histogram_plot(
  hcmst, 
  "relationship_quality", 
  "same_sex_couple", 
  plot_title = "Relationship Quality Frequency"
)

plot_grid(
  hist_sex_freq, 
  hist_flirt_freq, 
  hist_fight_freq, 
  hist_inc_change_freq,
  hist_sub_covid_freq,
  hist_par_covid_freq,
  hist_sub_vax_freq,
  hist_par_vax_freq,
  hist_cov_approach_freq,
  hist_quality_freq,
  ncol = 1)
```

It can be noted that the sample of `same_sex_couples` is significantly smaller that that of different-sex couples..

Same-sex and different-sex couples show similar patters in how frequently they engage in sexual activity. However, the overall sample is skewed towards reporting low sex frequency. We can also visualize how often couples flirt with or fight with their partners. The data indicates a skew towards lower frequencies for both flirting and fighting.

In the context of the COVID-19 pandemic, it can be noted that most subjects had no change in income during the pandemic, did not have COVID, were fully vaccinated including the booster shot, and completely agreed with their partners in terms of their approach to the pandemic.

As for our target variable, most participants perceived the quality of their relationship to be `good` or `excellent`.

```{r img_save, echo = FALSE}
ggsave("../img/hcmst.png", plot = hist_flirt_freq, width = 6, height = 4, dpi = 300)
```

#### 4. Model Fitting and Assumptions

We will be assuming and fitting a proportional odds model, which assumes that the log-odds of being in a higher category of the response variable follows a linear relationship with the explanatory variables.

For our modeling purposes, we will use the `MASS` package, `polr()` function to obtain model estimates.

Given that we are using ordinal categorical variables as explanatory variables, we will also set our model to use successive differences contrasts to facilitate the interpretation of the estimates.

```{r contrasts}
options(contrasts = c("contr.treatment", "contr.sdif"))
```

Now, we can fit our model. Noting that `polr()` does not calculate p-values, we can compute them and bind them to our model estimates. We will also focus only on statistically significant estimates.

For the purposes of this exercise, we will only use a subset of the variables in our modeling. We will be interested in how behavioral variables, as well as COVID-19 related variables, affect the perception of relationship quality. Additionally, we are interested in seeing if these effects are different for same sex couples.

```{r model_fit}
ordinal_model <- polr(
  relationship_quality ~ same_sex_couple + 
    sex_frequency + 
    flirts_with_partner + 
    fights_with_partner + 
    inc_change_during_pandemic + 
    subject_had_covid + 
    partner_had_covid + 
    subject_vaccinated + 
    partner_vaccinated + 
    agree_covid_approach,
  data = hcmst, Hess = TRUE
)
```

To ensure the validity of our analysis, we are going to test the proportional odds assumption. To do so, we can use the Brant-Wald test, which assesses whether our model fulfills this assumption.

```{r brant_test, results = "hide"}
bt <- brant(ordinal_model)
```

```{r table_a}
kable(bt, digits = 2)
```

Note that with an $\alpha = 0.05$, we don't have evidence of violating the proportional odds assumption (the column `probability` delivers the corresponding p-values).

Also, note that throughout the fitting process and the Brant-Wald test we got a few warnings. These are caused by having categories in our model that are quite sparse, where combinations of the response and the levels of the predictor have no observations. These gaps may affect Brant statistics. A solution for this matter would be to bin some of the levels of our categories to avoid having empty combinations of levels between our response and explanatory variables. For now, we will only acknowledge this issue.

#### 5. Model Interpretation

First, we will calculate the P-value (as `polr()`does not calculate it) and we are going to exponentiate the estimates for easier interpretation of the log-odds.

```{r model_int}
summary_partial_ordinal_model <- cbind(tidy(ordinal_model),
  p.value = pnorm(abs(tidy(ordinal_model)$statistic), lower.tail = FALSE) * 2) |>
  mutate_if(is.numeric, round, 2)

summary_partial_ordinal_model <- summary_partial_ordinal_model |>
  mutate(exp_estimate = exp(estimate)) |> 
  filter(p.value < 0.05, 
         coef.type == "coefficient")

kable(summary_partial_ordinal_model, digits = 2)
```

Positive estimates are associated with higher relationship quality, while negative estimates are associated with lower relationship quality. Additionally, the exponential of the estimate is the odds ratio, which represents how much the odds of having a higher `relationship_quality` level changes between levels.

We can interpret these estimates:

-   `same_sex_couple`: Being in a same-sex couple is associated with 41% lower odds of reporting better relationship quality compared to different-sex couples.
-   `sex_frequency`: People who have sex 2-3 times a month have 1.9 higher odds of reporting better relationship quality that those who have sex once a month or less.
-   `flirts_with_partner`: Flirting every day vs a few times per week is associated with 2.4 higher odds of better relationship quality.
-   `fights_with_partner`: Fighting once vs none is associated with 48% lower odds of better relationship quality.
-   `inc_change_during_pandemic`: Individuals whose income did not change during the pandemic had 1.7 times higher odds of reporting better relationship quality vs those whose income was worse.
-   `partner_vaccinated`: Having a partner who is partially vaccinated is associated with 2.4 times higher odds of having a better relationship quality.
-   `agree_covid_approach`: Mostly agreeing with a partner's approach to COVID vs mostly disagreeing is associated with 3.3 times higher odds of better relationship quality.
-   `agree_covid_approach`: Completely agreeing with a partner's approach to COVID vs mostly agreeing is associated with 1.8 times higher odds of better relationship quality.

#### 6. Prediction

With our fitted model, we are now able to make predictions on new examples. Based on behavioral and COVID-19 related variables, we can predict the probabilities of the different levels of perceived `relationship_quality` for a specific subject. Focusing on the levels that were associated with statistically significant factors:

```{r prediction}
round(predict(ordinal_model, tibble(
  same_sex_couple = "Yes",
  sex_frequency = "Once a Month or Less",
  flirts_with_partner = "A Few Times a Week",
  fights_with_partner = "One Time",
  inc_change_during_pandemic = "Worse",
  subject_had_covid = "No",
  partner_had_covid = "No", 
  subject_vaccinated = "Not Vaccinated", 
  partner_vaccinated = "Not Vaccinated",
  agree_covid_approach = "Mostly Disagree"
  ), 
  type = "probs"),2)

round(predict(ordinal_model, tibble(
  same_sex_couple = "No",
  sex_frequency = "Two or Three Times a Month",
  flirts_with_partner = "Every Day",
  fights_with_partner = "Zero Times",
  inc_change_during_pandemic = "No Change",
  subject_had_covid = "No",
  partner_had_covid = "No", 
  subject_vaccinated = "Not Vaccinated", 
  partner_vaccinated = "Partially Vaccinated",
  agree_covid_approach = "Mostly Agree"
  ), 
  type = "probs"), 2)
```

We can note that a subject involved in a same-sex couple, who has sex once a month or less, flirts with their partner a few times a week, fought with their partner once in the last week, experienced worsened income during the pandemic, is not vaccinated (nor is their partner), and mostly disagrees with their partner's approach to COVID, has a 41% probability of perceiving their relationship quality as `fair` and a combined probability of 40% of categorizing it as `poor` or `very poor`.

On the other hand, a subject involved in a different-sex couple, who has sex two or three times a month, flirts with their partner on a daily basis, has not fought with their partner in the last week, experienced no change in income during the pandemic, has a partially vaccinated partner, and mostly agrees with their partner's approach to COVID, has a 78% probability of perceiving their relationship quality as `excellent`, and a 98% probability of rating it as `good` or `excellent`.

### Discussion

This analysis explored how relationship habits, particularly in the context of the COVID-19 pandemic relate to perceived relationship quality across different types of couples. Through the analysis, results highlight that relationship behaviors and shared beliefs meaningfully affect how individuals evaluate their relationships. More specifically, we could conclude the following:

-   **Behaviors are important**: More frequent sex and flirting are associated with higher perceived relationship quality, while more frequent fighting affects that perception negatively.

-   **Shared beliefs are beneficial**: Agreement with a partner's approach to COVID-19 was a very strong predictor of relationship quality. Moving in a direction of agreement had a significant positive association.

-   **External factors play a role as well**: Financial stability during the pandemic, as well as the partner's vaccination status were also linked to better relationship quality, suggesting that alignment on external factors, such as health and financial concerns, could cause stress in couple dynamics contribute to positive perceptions.

-   **Same-sex couples appear to have lower quality in their relationships**: Contrary to expectations, being in a same-sex couple was significantly associated with lower perceived relationship quality. While the effect was not as strong as other factors, it was still significant and may warrant further exploration into factors contributing to this difference.

-   **Encoding and model choice are critical**: Model selection is tied to the type of response variable. Ordinal categorical response variables are tied to ordinal logistic regression models. Also, encoding categorical predictors must be done with care to ensure meaningful and interpretable contrasts.

-   **Sparsity may affect results**: When categorical variables have many levels, certain combinations of predictors may be underrepresented. This sparsity may impact the model, so caution is advised when drawing conclusions from underpopulated subgroups.
